\hypertarget{dictionaries}{%
\chapter{Dictionaries}\label{dictionaries}}

In the previous chapter we used a \passthrough{\lstinline!for!} loop to
read a file and count the words. In this chapter, you'll learn about a
new type called a \textbf{dictionary}, and we'll use it to count the
number of unique words and the number of times each one appears.

Along the way, you will also see how to use an index to select an
element from a sequence (tuple, list, or array). And you will learn a
little about Unicode, which is used to represent letters, numbers, and
punctuation for almost every language in the world.

\hypertarget{indexing}{%
\section{Indexing}\label{indexing}}

Suppose you have a variable named \passthrough{\lstinline!t!} that
refers to a list or tuple. You can select an element using the bracket
operator, \passthrough{\lstinline![]!}. For example, here's a tuple of
strings:

\begin{lstlisting}[]
t = 'zero', 'one', 'two'
\end{lstlisting}

To select the first element, we put \passthrough{\lstinline!0!} in
brackets:

\begin{lstlisting}[]
t[0]
(@\dashfill@)
@@@'zero'@@@
\end{lstlisting}

To select the second element, we put \passthrough{\lstinline!1!} in
brackets:

\begin{lstlisting}[]
t[1]
(@\dashfill@)
@@@'one'@@@
\end{lstlisting}

To select the third element, we put \passthrough{\lstinline!2!} in
brackets:

\begin{lstlisting}[]
t[2]
(@\dashfill@)
@@@'two'@@@
\end{lstlisting}

The number in brackets is called an \textbf{index} because it indicates
which element we want. Tuples and lists use zero-based numbering; that
is, the index of the first element is 0. Some other programming
languages use one-based numbering. There are pros and cons of both
systems (see \url{https://en.wikipedia.org/wiki/Zero-based_numbering}).

The index in brackets can also be a variable:

\begin{lstlisting}[]
i = 1
t[i]
(@\dashfill@)
@@@'one'@@@
\end{lstlisting}

Or an expression with variables, values, and operators:

\begin{lstlisting}[]
t[i+1]
(@\dashfill@)
@@@'two'@@@
\end{lstlisting}

But if the index goes past the end of the list or tuple, you get an
error.

Also, the index has to be an integer; if it is any other type, you get
an error.

\textbf{Exercise:} You can use negative integers as indices. Try using
\passthrough{\lstinline!-1!} and \passthrough{\lstinline!-2!} as
indices, and see if you can figure out what they do.

\hypertarget{dictionaries-1}{%
\section{Dictionaries}\label{dictionaries-1}}

A dictionary is similar to a tuple or list, but in a dictionary, the
index can be almost any type, not just an integer. We can create an
empty dictionary like this:

\begin{lstlisting}[]
d = {}
\end{lstlisting}

Then we can add elements like this:

\begin{lstlisting}[]
d['one'] = 1
d['two'] = 2
\end{lstlisting}

In this example, the indices are the strings,
\passthrough{\lstinline!'one'!} and \passthrough{\lstinline!'two'!} If
you display the dictionary, it shows each index and the corresponding
value.

\begin{lstlisting}[]
d
(@\dashfill@)
@@@{'one': 1, 'two': 2}@@@
\end{lstlisting}

Instead of creating an empty dictionary and then adding elements, you
can create a dictionary and specify the elements at the same time:

\begin{lstlisting}[]
d = {'one': 1, 'two': 2, 'three': 3}
d
(@\dashfill@)
@@@{'one': 1, 'two': 2, 'three': 3}@@@
\end{lstlisting}

When we are talking about dictionaries, an index is usually called a
\textbf{key}. In this example, the keys are strings and the
corresponding values are integers.

A dictionary is also called a \textbf{map}, because it represents
correspondence or ``mapping'', between keys and values. So we might say
that this dictionary maps from English number names to the corresponding
integers.

You can use the bracket operator to select an element from a dictionary,
like this:

\begin{lstlisting}[]
d['two']
(@\dashfill@)
@@@2@@@
\end{lstlisting}

But don't forget the quotation marks. Without them, Python looks for a
variable named \passthrough{\lstinline!two!} and doesn't find one.

To check whether a particular key is in a dictionary, you can use the
special word \passthrough{\lstinline!in!}:

\begin{lstlisting}[]
'one' in d
(@\dashfill@)
@@@True@@@
\end{lstlisting}

\begin{lstlisting}[]
'zero' in d
(@\dashfill@)
@@@False@@@
\end{lstlisting}

The word \passthrough{\lstinline!in!} is actually an operator in Python,
so you can't use it as a variable name.

If a key is already in a dictionary, adding it again has no effect:

\begin{lstlisting}[]
d
(@\dashfill@)
@@@{'one': 1, 'two': 2, 'three': 3}@@@
\end{lstlisting}

\begin{lstlisting}[]
d['one'] = 1
d
(@\dashfill@)
@@@{'one': 1, 'two': 2, 'three': 3}@@@
\end{lstlisting}

But you can change the value associated with a key:

\begin{lstlisting}[]
d['one'] = 100
d
(@\dashfill@)
@@@{'one': 100, 'two': 2, 'three': 3}@@@
\end{lstlisting}

You can loop through the keys in a dictionary like this:

\begin{lstlisting}[]
for key in d:
    print(key)
(@\dashfill@)
@@@one
two
three@@@
\end{lstlisting}

If you want the keys and the values, one way to get them is to loop
through the keys and look up the values:

\begin{lstlisting}[]
for key in d:
    print(key, d[key])
(@\dashfill@)
@@@one 100
two 2
three 3@@@
\end{lstlisting}

Or you can loop through both at the same time, like this:

\begin{lstlisting}[]
for key, value in d.items():
    print(key, value)
(@\dashfill@)
@@@one 100
two 2
three 3@@@
\end{lstlisting}

The \passthrough{\lstinline!items!} method loops through the key-value
pairs in the dictionary; each time through the loop, they are assigned
to \passthrough{\lstinline!key!} and \passthrough{\lstinline!value!}.

\textbf{Exercise:} Make a dictionary with the numbers
\passthrough{\lstinline!1!}, \passthrough{\lstinline!2!}, and
\passthrough{\lstinline!3!} as keys and strings as values. The strings
should be the numbers ``one'', ``two'', and ``three'' in any language
you know.

Write a loop that prints just the values from the dictionary.

\hypertarget{unique-words}{%
\section{Unique Words}\label{unique-words}}

In the previous chapter we downloaded \emph{War and Peace} from Project
Gutenberg and counted the number of lines and words. Now that we have
dictionaries, we can also count the number of unique words and the
number of times each one appears.

As we did in the previous chapter, we can read the text of \emph{War and
Peace} and count the number of words.

\begin{lstlisting}[]
fp = open('2600-0.txt')
count = 0
for line in fp:
    count += len(line.split())
    
count
(@\dashfill@)
@@@566317@@@
\end{lstlisting}

To count the number of unique words, I'll loop through the words in each
line and add them as keys in a dictionary:

\begin{lstlisting}[]
fp = open('2600-0.txt')
unique_words = {}
for line in fp:
    for word in line.split():
        unique_words[word] = 1
\end{lstlisting}

This is the first example we've seen with one loop inside another.

\begin{itemize}
\item
  The outer loop runs through the lines in the file.
\item
  The inner loops runs through the words in each line.
\end{itemize}

Each time through the inner loop, we add a word as a key in the
dictionary, with the value 1. If the same word appears more than once,
it gets added to the dictionary again, which has no effect. So the
dictionary contains only one copy of each unique word in the file.

At the end of the loop, we can display the first 10 keys:

\begin{lstlisting}[]
i = 0
for key in unique_words:
    print(key)
    i += 1
    if i == 10:
        break
(@\dashfill@)
@@@
The
Project
Gutenberg
EBook
of
War
and
Peace,
by@@@
\end{lstlisting}

The dictionary contains all the words in the file, in order of first
appearance. But each word only appears once, so the number of keys is
the number of unique words:

\begin{lstlisting}[]
len(unique_words)
(@\dashfill@)
@@@41991@@@
\end{lstlisting}

It looks like there are about 42,000 different words in the book, which
is substantially less than the total number of words, about 560,000. But
that's not quite right, because we have not taken into account
capitalization and punctuation.

\textbf{Exercise:} Before we deal with that problem, let's practice with
\textbf{nested loops}, that is, one loop inside another. Suppose you
have a list of words, like this:

\begin{lstlisting}[]
line = ['War', 'and', 'Peace']
\end{lstlisting}

Write a nested loop that iterates through each word in the list, and
each letter in each word, and prints the letters on separate lines.

\hypertarget{capitalization}{%
\section{Capitalization}\label{capitalization}}

When we count unique words, we probably want to treat
\passthrough{\lstinline!The!} and \passthrough{\lstinline!the!} as the
same word. We can do that by converting all words to lower case, using
the \passthrough{\lstinline!lower!} function:

\begin{lstlisting}[]
word = 'The'
word.lower()
(@\dashfill@)
@@@'the'@@@
\end{lstlisting}

\passthrough{\lstinline!lower!} creates a new string; it does not modify
the original string.

\begin{lstlisting}[]
word
(@\dashfill@)
@@@'The'@@@
\end{lstlisting}

However, you can assign the new string back to the existing variable,
like this:

\begin{lstlisting}[]
word = word.lower()
\end{lstlisting}

Now if we can display the new value of \passthrough{\lstinline!word!},
we get the lowercase version:

\begin{lstlisting}[]
word
(@\dashfill@)
@@@'the'@@@
\end{lstlisting}

\textbf{Exercise:} Modify the previous loop so it makes a lowercase
version of each word before adding it to the dictionary. How many unique
words are there, if we ignore the difference between uppercase and
lowercase?

\hypertarget{punctuation}{%
\section{Punctuation}\label{punctuation}}

To remove punctuation from the words, we can use
\passthrough{\lstinline!strip!}, which removes specified characters from
the beginning and end of a string. Here's an example:

\begin{lstlisting}[]
word = 'abracadabra'
word.strip('ab')
(@\dashfill@)
@@@'racadabr'@@@
\end{lstlisting}

In this example, \passthrough{\lstinline!strip!} removes all instances
of \passthrough{\lstinline!a!} and \passthrough{\lstinline!b!} from the
beginning and end of the word, but not from the middle. But note that it
makes a new word; it doesn't modify the original:

\begin{lstlisting}[]
word
(@\dashfill@)
@@@'abracadabra'@@@
\end{lstlisting}

To remove punctuation, we can use the \passthrough{\lstinline!string!}
library, which provides a variable named
\passthrough{\lstinline!punctuation!}.

\begin{lstlisting}[]
import string

string.punctuation
\end{lstlisting}

\begin{lstlisting}
'!"#$%&\'()*+,-./:;<=>?@[\\]^_`{|}~'
\end{lstlisting}

\passthrough{\lstinline!string.punctuation!} contains the most common
punctuation marks, but as we'll see, not all of them. Nevertheless, we
can use it to handle most cases. Here's an example:

\begin{lstlisting}[]
line = "It's not given to people to judge what's right or wrong."

for word in line.split():
    word = word.strip(string.punctuation)
    print(word)
(@\dashfill@)
@@@It's
not
given
to
people
to
judge
what's
right
or
wrong@@@
\end{lstlisting}

\passthrough{\lstinline!strip!} removes the period at the end of
\passthrough{\lstinline!wrong!}, but not the apostrophes in
\passthrough{\lstinline!It's!}, \passthrough{\lstinline!don't!} and
\passthrough{\lstinline!what's!}. So that's good, but we have one more
problem to solve. Here's another line from the book.

\begin{lstlisting}[]
line = 'anyone, and so you don’t deserve to have them.”'
\end{lstlisting}

Here's what happens when we try to remove the punctuation.

\begin{lstlisting}[]
for word in line.split():
    word = word.strip(string.punctuation)
    print(word)
(@\dashfill@)
@@@anyone
and
so
you
don’t
deserve
to
have
them.”@@@
\end{lstlisting}

It removes the comma after \passthrough{\lstinline!anyone!}, but not the
period and quotation mark after \passthrough{\lstinline!them!}. The
problem is that this kind of quotation mark is not in
\passthrough{\lstinline!string.punctuation!}.

To fix this problem, we'll use the following loop, which

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Reads the file and builds a dictionary that contains all punctuation
  marks that appear in the book, then
\item
  It uses the \passthrough{\lstinline!join!} function to concatenate the
  keys of the dictionary in a single string.
\end{enumerate}

You don't have to understand everything about how it works, but you
should read it and see how much you can figure out. You can read the
documentation of the \passthrough{\lstinline!unicodedata!} library here
at \url{https://docs.python.org/3/library/unicodedata.html}.

\begin{lstlisting}[]
import unicodedata

fp = open('2600-0.txt')
punc_marks = {}
for line in fp:
    for x in line:
        category = unicodedata.category(x)
        if category[0] == 'P':
            punc_marks[x] = 1
        
all_punctuation = ''.join(punc_marks)
print(all_punctuation)
(@\dashfill@)
@@@,.-:[#]*/“’—‘!?”;()%@@@@
\end{lstlisting}

\textbf{Exercise:} Modify the word-counting loop from the previous
section to convert words to lower case \emph{and} strip punctuation
before adding them to the dictionary. Now how many unique words are
there?

Optional: You might want to skip over the front matter and start with
the text of Chapter 1, and skip over the license at the end, as we did
in the previous chapter.

\hypertarget{word-frequencies}{%
\section{Word Frequencies}\label{word-frequencies}}

In the previous section we counted the number of unique words, but we
might also want to know how often each word appears. Then we can find
the most common and least common words in the book. To count the
frequency of each word, we'll make a dictionary that maps from each word
to the number of times it appears.

Here's an example that loops through a string and counts the number of
times each letter appears.

\begin{lstlisting}[]
word = 'Mississippi'

letter_counts = {}
for x in word:
    if x in letter_counts:
        letter_counts[x] += 1
    else:
        letter_counts[x] = 1
        
letter_counts
(@\dashfill@)
@@@{'M': 1, 'i': 4, 's': 4, 'p': 2}@@@
\end{lstlisting}

The \passthrough{\lstinline!if!} statement here uses a feature we have
not seen before, an \passthrough{\lstinline!else!} clause. Here's how it
works.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  First, it checks whether the letter, \passthrough{\lstinline!x!}, is
  already a key in the dictionary,
  \passthrough{\lstinline!letter\_counts!}.
\item
  If so, it runs the first statement,
  \passthrough{\lstinline!letter\_counts[x] += 1!}, which increments the
  value associated with the letter.
\item
  Otherwise, it runs the second statement,
  \passthrough{\lstinline!letter\_counts[x] = 1!}, which adds
  \passthrough{\lstinline!x!} as a new key, with the value
  \passthrough{\lstinline!1!} indicating that we have seen the new
  letter once.
\end{enumerate}

The result is a dictionary that maps from each letter to the number of
times it appears.

To get the most common letters, we can use a
\passthrough{\lstinline!Counter!}, which is similar to a dictionary. To
use it, we have to import a library called
\passthrough{\lstinline!collections!}:

\begin{lstlisting}[]
import collections
\end{lstlisting}

Then we use \passthrough{\lstinline!collections.Counter!} to convert the
dictionary to a \passthrough{\lstinline!Counter!}:

\begin{lstlisting}[]
counter = collections.Counter(letter_counts)
type(counter)
(@\dashfill@)
@@@collections.Counter@@@
\end{lstlisting}

\passthrough{\lstinline!Counter!} provides a function called
\passthrough{\lstinline!most\_common!} we can use to get the most common
characters:

\begin{lstlisting}[]
counter.most_common(3)
(@\dashfill@)
@@@[('i', 4), ('s', 4), ('p', 2)]@@@
\end{lstlisting}

The result is a list of tuples, where each tuple contains a character
and an integer.

\textbf{Exercise:} Modify the loop from the previous exercise to count
the frequency of the words in \emph{War and Peace}; then print the 20
most common words and the number of times each one appears.

\textbf{Exercise:} You can run \passthrough{\lstinline!most\_common!}
with no value in parentheses, like this:

\begin{lstlisting}
word_freq_pairs = counter.most_common()
\end{lstlisting}

The result is a list of tuples, with one tuple for every unique word in
the book. Assign the result to a variable so it doesn't get displayed.
Then answer the following questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  How many times does the \#1 ranked word appear (that is, the first
  element of the list)?
\item
  How many times does the \#10 ranked word appear?
\item
  How many times does the \#100 ranked word appear?
\item
  How many times does the \#1000 ranked word appear?
\item
  How many times does the \#10000 ranked word appear?
\end{enumerate}

Do you see a pattern in the results? We will explore this pattern more
in the next chapter.

\textbf{Exercise:} Write a loop that counts how many words appear 200
times. What are they? How many words appear 100 times, 50 times, and 20
times?

\textbf{Optional:} If you know how to define a function, write a
function that takes a \passthrough{\lstinline!Counter!} and a frequency
as arguments, prints all words with that frequency, and returns the
number of words with that frequency.

\hypertarget{summary}{%
\section{Summary}\label{summary}}

